Microclimatic spatial variation buffers local warming in alpine vegetation

### 2022-11-21 doc/manuscript.Rmd

The study was conducted in the central calcareous massif of the Picos de Europa National Park, in northern Spain (**Figure 1A**). The study area is a biodiversity hotspot for cold-adapted plants in the Iberian Peninsula and a biogeographical hub for Alpine and Mediterranean lineages in Western Europe (XXX). The central calcareous massif occupies c. 50 km^2^ and supports a high diversity of ecosystems, with alpine vegetation mostly occurring between 1900 and 2400 m a.s.l., with a local species pool of XX species (Jiménez-Alfaro et al. XXX). In 2008, we established a long-term ecological research program for monitoring soil climate and vegetation change. We selected four study sites along a north-south gradient (**Figure 1B**), reflecting the main macroclimatic variation from Atlantic to Mediterranean influence. The sites were selected to... general design and description of sites... (Borja)

@@

*Spatial survey.* In 2018, we visited the same areas to study the spatial variation of vegetation and microclimate around the previously sampled areas. Using the long-term temperature logger as the central point, we additionally placed 20 iButtons (Thermochron, iButton, Newbury, UK; accuracy: +/- 0.5 ºC from -10 ºC to +65 ºC, resolution: 0.5 ºC, records each 4 hours) in 20 plots of 1 m^2^ separated 10 m from each other in the four cardinal directions (**Figure 1C**). At each of the 20 plots, we identified vascular plants and estimated their relative cover in %. In 2019 we came back to download the data of the iButtons. These data, together with the vegetation data of the iButton plots, is the "spatial survey".

@@

To construct scenarios of climatic changes, we used the temporal survey data (10 years, 2009-2018) to describe situations in which the extreme values of today would become the new average. To do so, we calculated, for GDD and FDD, the maximum and minimum values recorded in the entire period. With the resulting four values, we created four scenarios: hot and snowy (max GDD, min FDD), hot and frozen (max GDD, max FDD), cold and snowy (min GDD, min FDD) and cold and frozen (min GDD, max FDD). We used the GLM models to predict the probably of presence for each species and scenario, considering that a probability of 0 in a given scenario would mean the extinction of the species in said scenario.

@@

Along the 10 years of soil temperature monitoring, we found a general trend of increasing temperature, especially in the southwest and warmest site (Hoyo sin Tierra) (Figure 3). Two of the sites (Los Cazadores and Los Boches) showed a consistent pattern of continuous snow cover during winter (reflected by temperature records around 0º C). In contrast, the two other sites (Hoy Sin Tierri and Hoyo Sin Tierra) showed frost temperatures during most winters (although they may not be constant and intermixed with short snow periods). Maximum temperatures were... Soil temperature corresponding to the spatial surveys also showed high variation within sites (Figure 3). In the four sites we found plots with contrasting patterns of snow cover and frost during winter, reflected by... When looking at the whole microclimatic variation among plots (Figure 4), we found that the four sites were represented by a wide range of temperatures, with the exception of one site (Los Cazadores). The two main axes of variation were mainly related to FDD and GDD, the latter also correlated with.... (Eduardo)

## Plant diversity and microclimate

Across the whole study system (temporal and spatial surveys) we recorded 86 species of vascular plants (considering *Helianthemum apenninum* subsp. *urrielense* and *Helianthemum apenninum* subsp. *cantabricum* as separate species), representing % of the total species pool of the study region. The five most frequent species were *Thymus praecox* subsp. *ligusticus* (83 occurences), *Anthyllis vulneraria* (73), *Koeleria vallesiana* (59), *Minuartia verna* (55) and *Helianthemum canum* (52). Average species richness per 1m^2^ plot was 13, with the richest plot having 25 species and the poorest one just two.

@@

Changes in species frequency were/were not correlated with XX and XX.... (Table 1) (Eduardo). In the spatial survey (20 plots per site, n = 80; including one plot placed in rocks with no vascular plants) we recorded 81 species, 36 of which were absent from the temporal survey plots.

Changes in species composition (**Fig. 4**) were associated to GDD (NMDS axis 1) to represent a main gradient from the warmest to the coldest microsites, and to FDD (NMDS2) in a gradient of snow versus frost winters. In the compositional space, the sites were poorly differentiated across the four sites (PERMANOVA, adonis) mainly by the effect of one site (Los Boches) with most plots places in the cold and frost margins. Etc.... (Borja).

redo environmental fit

Results of the environmental fit:

```{r}
print(ef1)
```

Correlations between bioclimatic variables:

```{r}
print(biocor)
```

Let us keep GDD, bio7 and FDD for plotting NMDS (low autocorrelation, high r2 in environmental fit)

## Microclimatic ranges in time and space

.... (Eduardo) (quizás repensar esta parte, pero es importante mantenerla porque es la clave).

>>>>

The study was conducted in the central calcareous massif of the Picos de Europa National Park, in northern Spain (**Fig. 1A**). The study area is a biodiversity hotspot for cold-adapted plants in the Iberian Peninsula and a biogeographical hub for Alpine and Mediterranean lineages in Western Europe (XXX). The central calcareous massif occupies c. 50 km^2^ and supports a high diversity of ecosystems, with alpine vegetation mostly occurring between 1900 and 2400 m a.s.l., with a local species pool of XX species (Jiménez-Alfaro et al. XXX). In 2008, we established a long-term ecological research program for monitoring soil climate and vegetation change. We selected four study sites along a north-south gradient (**Fig. 1B**), reflecting the main macroclimatic variation from Atlantic to Mediterranean influence. The sites were selected to... general design and description of sites... (Borja)

@@

*Spatial survey.* In 2018, we visited the same areas to study the spatial variation of vegetation and microclimate around the previously sampled areas. Using the long-term temperature logger as the central point, we additionally placed 20 iButtons (Thermochron, iButton, Newbury, UK; accuracy: +/- 0.5 ºC from -10 ºC to +65 ºC, resolution: 0.5 ºC, records each 4 hours) in 20 plots of 1 m^2^ separated 10 m from each other in the four cardinal directions (**Fig. 1C**). At each of the 20 plots, we identified vascular plants and estimated their relative cover in %. In 2019 we came back to download the data of the iButtons. These data, together with the vegetation data of the iButton plots, is the "spatial survey".

@@

To construct scenarios of climatic changes, we used the temporal survey data (10 years, 2009-2018) to describe situations in which the extreme values of today would become the new average. To do so, we calculated, for GDD and FDD, the maximum and minimum values recorded in the entire period. With the resulting four values, we created four scenarios: hot and snowy (max GDD = 2069 ºC, min FDD = 0 ºC), hot and frozen (max GDD = 2069 ºC, max FDD = 247 ºC), cold and snowy (min GDD = 570 ºC, min FDD = 0 ºC) and cold and frozen (min GDD = 570 ºC, max FDD = 247 ºC). We used the GLM models to predict the probably of presence for each species and scenario, considering that a probability of 0 in a given scenario would mean the extinction of the species in said scenario.

@@

After 10 years of soil temperature monitoring (**Fig. 2A**), we found that two of the sites (Los Cazadores and Los Boches) showed a consistent pattern of continuous snow cover during winter (i.e. snowbed conditions reflected by temperature records around 0 ºC). In contrast, the two other sites (Ḥou Sin Tierri and Hoyo Sin Tierra) showed frost temperatures during most winters (i.e. fellfield conditions). The length of snow cover went from 72 days at Hoyo Sin Tierra to 193 days at Los Boches. Mean annual temperature ranged from 3.8 ºC (Los Boches) to 6.2 ºC (Hoyo Sin Tierra). The average annual range went from 19.4 ºC (Los Boches) to 23.8 ºC (Hoyo Sin Tierra), and the diurnal range from 2 ºC (Los Boches) to 3 ºC (Ḥou Sin Tierri). The absolute maximum was 27.4 ºC, reached at Hoyo Sin Tierra in July 2015, while the absolute minimum was -12 ºC at Ḥou Sin Tierri in December 2016. Average annual GDD were lowest at Los Boches (930 ºC) and highest at Hoyo Sin Tierra (1,711 ºC). FDD were low at Los Cazadores (3 ºC) and Los Boches (26 ºC), and high at Ḥou Sin Tierri (89 ºC) and Hoyo Sin Tierra (100 ºC).

Soil temperature corresponding to the spatial surveys (**Fig. 2B**) showed high variation within sites. In the four sites we found plots representing both snowbed and fellfield conditions. Within the plots of a same site, the length of snow cover could go from non-existent (i.e. 0 days under snow and frost temperatures during the whole year) to as long as 8 months (with a maximum of 234 days, from November to July). The annual range of temperatures went from 17.8 ºC to 30.3 ºC, and the diurnal range from 1.6 ºC to 5.5 ºC. The absolute maximum was 33 ºC, the absolute minimum -12 ºC. GDD ranged from 517 ºC to 1,612 ºC, and FDD from 0 ºC to 206 ºC.

To compare the width of the microclimatic variability in time and space we took, for each site, the difference between the maximum and minimums values of GDD and FDD in time (maxi and min values in 10 years per site) and space (max and min values in 20 plots per site) (**Fig. 3A**). For GDD, the spatial difference was larger than the temporal one, but the difference was only marginally different (paired t-test, one.sided, t = 1.945, df = 3, p-value = 0.073). For FDD, the two differences were not significantly different (paired t-test, one.sided, t = -0.29029, df = 3, p-value = 0.6047).

> Quizás sería mejor quitar este último párrafo y la figura 3, y centrarse en los modelos, la parte de time vs. space no es nada concluyente. Quizás describirla mejor, dentro del estilo de los dos párrafos anteriores.

## Plant diversity

Across the whole study system (temporal and spatial surveys) we recorded 86 species of vascular plants (considering *Helianthemum apenninum* subsp. *urrielense* and *Helianthemum apenninum* subsp. *cantabricum* as separate species), representing % of the total species pool of the study region. Of these, 81 species were in the spatial survey plots, and 48 in the temporal survey plots. The five most frequent species were *Thymus praecox* subsp. *ligusticus* (83 occurences), *Anthyllis vulneraria* (73), *Koeleria vallesiana* (59), *Minuartia verna* (55) and *Helianthemum canum* (52). Average species richness per 1m^2^ plot was 13, with the richest plot having 25 species and the poorest one just two.

@@

## Species responses to microclimate

Changes in species composition (**Fig. 4**) were associated to GDD (NMDS axis 1, p < 0.001, R^2^ = 0.778) to represent a main gradient from the warmest to the coldest microsites, and to FDD (NMDS2, p < 0.001, R^2^ = 0.330) in a gradient of snow versus frost microsites In the compositional space, the sites were poorly differentiated along the first axis, although Los Boches occupied the cold and frozen space.

Of the 81 species in the spatial plots, 36 had more than 10 occurrences, and we included them in the GLM modeling (full model results in supplementary material **S3**). For 16 of these species we produced models that we considered satisfactory (i.e. at least one of the two bioclimatic indices had a significant effect size and the value of McFadden's pseudo R^2^ was higher than 0.15) (**Table 1**). We found that some species survived only in the hot scenarios (e.g. *Androsace villosa*), the cold scenario (e.g. *Festuca glacialis*) or the snowy scenarios (e.g. *Alchemilla catalaunica*). The cold & snowy scenario produced the lowest number of species extinctions and the hot & frozen scenario the most extinctions, with the cold & frozen and the hot & snowy scenarios producing intermediate numbers of extinctions (**Fig. 5**).
